---
title: "Class 09: Candy Mini Project"
author: "Mason Dinh (PID: A19140455)"
format: pdf
toc: true
---

## Background

Today we are taking a small detour to analyze a dataset (that we have more intrinsic insight into) with the most useful analysis method we have learned thus far - Principal Component Analysis (PCA)

## Data Import
The data is all related to halloween candy is from the 538 website.

```{r}
candy<-read.csv("candy-data.txt", row.names = 1)
head(candy)
```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```
There are 85 different types of candy in this dataset.


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```
There are 38 fruity candy types in this dataset.

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
library(dplyr)

candy |> 
  filter(row.names(candy)=="Air Heads") |> 
  select(winpercent, pricepercent, sugarpercent)
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent

```

There is a useful `skim()` function in the **skimr** package that can help give you a quick overview.

```{r}
library("skimr")
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

There are two columns that work differently than the other columns: The Histogram and Complete_Rate column look to be on a different scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

I think a one or zero represents if the candy contains choclate or not. 

## Exploratory Analysis

> Q8. Plot a histogram of winpercent values using both base R and ggplot2.

```{r}
hist(candy$winpercent)
```

```{r}
library(ggplot2)
ggplot(candy)+aes(winpercent)+geom_histogram(bins = 20,col="blue",fill="green")
```

> Q9. Is the distribution of winpercent values symmetrical?

No, the distribution is not symmetrical. It appears to be right skewed. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
summary(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean(candy$winpercent[candy$chocolate == 1])
mean(candy$winpercent[candy$fruity == 1])
```

> Q12. Is this difference statistically significant?

```{r}
t.test(winpercent ~ chocolate, data = candy)
```

Yes, the difference is statistically significant because the p-value is much less than 0.05. 

## Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
candy |> arrange(winpercent) |> select(winpercent)|> head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy |> arrange(desc(winpercent)) |> select(winpercent) |> head(5)
```

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r, fig.height=15,fig.width=10}
ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r, fig.height=15,fig.width=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
my_cols = rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r, fig.height=15,fig.width=10}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) +
  ylab("")
```

> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.


## Taking a look at pricepercent

```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent,pricepercent , label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miniatures appaer to offer the most bang for your buck.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip, Nestle Smarties, Ring Pop, Hershey's Krackel, and Hershey's Milk Chocolate	appear to be the most expensive candys in this data set. The least popular one being Nik L Nip.

```{r}
candy |> arrange(desc(pricepercent)) |> select(pricepercent) |> head(5)

```

## Exploring the correlation structure

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and Fruity are the two variables that are the most anti-correlated. 

> Q23. Use your corrplot result to make a prediction: which variables do you expect will have the largest contributions (i.e. loadings) to PC1 (i.e., drive the most separation between candies along the first principal component)?

From the plot, I think chocolate and fruity will have the largest contritbuion because they're the most anti-correlated variables on the plot and have the most negative values. 

## Principal Component Analysis

In this case we want to be sure to set `scale=TRUE()` argument for `prcomp()` because we have one variable, `winpercent()`, that is on a very different scale than all others and would otherwise dominate our PCA results.

```{r}
pca <- prcomp(candy, scale = TRUE)
summary(pca)
```

First major result figure in the "score plot" of PC1 vs PC2  - how different candy are related to each other on our new PC axis:

```{r}
plot(pca$x[, 1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r, eval=FALSE}
library(plotly) # ruined my PDF so I hid the result.
ggplotly(p)
```

```{r}
ggplot(pca$rotation) +
  aes(PC1, reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

The variables fruity, pluribus, and hard are picked up strongly by PC1 in the positive direction. We saw these relationships in the corrplot previously where these variables have positive correlations. 

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Based on the results of this lab, a "winning" candy's characteristics appear to be a chocolate bar with a higher price point, as there was correlation between the price and the popularity. Additonal characteristics like peanut-almondy or crisped rice wafer also appear to contribute to a winning candy. The barplot visualization allowed me to see what types of candy were the most popular, which was Reese's chocolate. The correlation helped me identify which characteristics are positively correlated with winpercent. Then lastly, the PCA showed that chocolate bar candy with additonal features were in the same direction as winpercent, which confirmed which characteristics make up a winning candy.
