---
title: "Class 12: Genome Informatics"
author: "Mason Dinh (PID: A19140455)"
format: pdf
toc: true
---

## Data Import

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

```{r}
dataset <- "https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt"
expr <- read.table(url(dataset))
```


```{r}
table(expr$geno)
```

```{r}
tapply(expr$exp, expr$geno, median)
```

> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
p <- boxplot(exp ~ geno, data=expr, 
             main="Gene Expression by Genotype",
             xlab="Genotype (rs8067378)", 
             ylab="Expression Level",
             col=c("lightblue", "lightgreen", "lightpink"))
```

From the plot, we can see that the A/A genotype has the highest median at 31.25 and the G/G genotype has the lowest at 20.07. Yes, the SNP rs8067378 affects the expression of ORMDL3, as seen from the boxplots. 

```{r}
p$stats
```






