---
title: "Class 13: Transcriptomics and the analysis of RNA-Seq data"
author: "Mason Dinh (PID: A19140455)"
format: pdf
toc: true
---

## Import Data

```{r}
library(BiocManager)
library(DESeq2)
```

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
head(counts)
```

```{r}
head(metadata)
```

> Q1. How many genes are in this dataset? 

There are `r nrow(counts)` genes in this dataset

> Q2. How many ‘control’ cell lines do we have? 

```{r}
sum(metadata$dex == "control")
```

## Toy differential gene expression

```{r}
colnames(counts) == metadata$id
```

```{r}
ncol(counts)
```

```{r}
metadata
```

- Find the "Control" columns in our `counts` object
- Extract just the "control" column values for all genes
- Calculate the average value per gene in these "control" columns

```{r}
control.inds <- metadata$dex == "control"
control.counts <- counts[ ,control.inds ]
control.mean <- rowMeans(control.counts)
```

> Q3. How would you make the above code in either approach more robust? Is there a function that could help here? 

To make the code above more robust, the control mean assumes that the metadata$dex is perfectly aligned with counts, so if a part of the sample is out of order or missing, then the calculated mean would be wrong. The more robust solution would be to use the all() command, which checks if every vector is TRUE. 

> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
treated.inds <- metadata$dex == "treated"
treated.counts <- counts[ ,treated.inds ]
treated.means <- rowMeans(treated.counts)
```

```{r}
treated.mean <- rowMeans(counts[, metadata$dex == "treated"])
```

For book-keeping, let's store these together as a new object called `meancounts`

```{r}
meancounts <- data.frame(control.mean, treated.mean)
head(meancounts)
```

> Q5 (a). Create a scatter plot showing the mean of the treated samples against the mean of the control samples.

```{r}
plot(meancounts[,1],meancounts[,2], xlab="Control", ylab="Treated")
```

> Q5 (b).You could also use the ggplot2 package to make this figure producing the plot below. What geom_?() function would you use for this plot? 

You would use the geom_point() function to make this type of plot with ggplot2. 

> Q6. Try plotting both axes on a log scale. What is the argument to plot() that allows you to do this? 

```{r}
plot(meancounts[,1], meancounts[,2], log="xy", xlab="Control", ylab="Treated")
```

We must often use log2 transform for this kind of data in bioinformatics because it makes my brain hurt less. 

```{r}
# Treated / Control

log2(20/20)
log2(40/20)
log2(10/20)
log2(80/20)
```

We call this little fraction the "log2 fold change" as it tells us how much more or less gene expression we have in units of doubling, etc. 

Lets calculate the log2 fold change for our `treated.mean` and `control.mean` counts and call this `log2fc`.

```{r}
meancounts$log2fc <- log2(meancounts$treated.mean / meancounts$control.mean)
head(meancounts)
```

A common "rule of thumb" threshold for calling a gene "up regulated" or "down regulated" is a log2 fold-change value of +2 or -2 (or greater)

> Q7. What is the purpose of the arr.ind argument in the which() function call above? Why would we then take the first column of the output and need to call the unique() function?

The arr.ind argument determines how the `which()` function reports the location of values that meet the criteria of `==0`.

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
sum(up.ind)
sum(down.ind)
```

> Q8. Using the up.ind vector above can you determine how many up regulated genes we have at the greater than 2 fc level? 

There are 250 up regulated genes that are greater than 2 fc level.

> Q9. Using the down.ind vector above can you determine how many down regulated genes we have at the greater than 2 fc level? 

There are 367 down regulated genes that are greated than 2 fc level.

> Q10. Do you trust these results? Why or why not?

No, these results should not be trusted because they do not determine whether the differences between groups is statistically significant, meaning that these results could be due to a lot of other variables or random noise.

## DESeq Analysis

Let's do this analysis properly and not forget about the significance of the differences. 

For this we will use the **DESeq2** package

```{r, message=FALSE}
library(DESeq2)
```

To run a DESeq analysis we need at least two inputs:

- `countData` i.e. are gene counts across different experiments
- `colData` i.e. our metadata about those count columns.

```{r}
dds<- DESeqDataSetFromMatrix(countData = counts, colData = metadata, design = ~dex)
```

Now we can run DESeq analysis pipeline using this `dds` object that has 

```{r}
dds<-DESeq(dds)
res<-results(dds)
head(res)
```

## Volcano plot 

This is a ubiquitous and common visualization for this type of data that puts the log2 foldchange and the adjusted p-value together in one plot that people can get insight for what is going on in the whole dataset results.

```{r}
library(ggplot2)
```

```{r}
ggplot(res) +
  aes(log2FoldChange, log(padj)) +geom_point(alpha=0.3)
```

That plot is not very useful because we don't care about genes with high p-values, we want the very low values below our alpha threshold.

```{r}
ggplot(res) +
  aes(log2FoldChange, -log(padj)) +geom_point(alpha=0.3)
```

> Q. Add annotations to this volcano plot including the log2 fold-change threshold of +2 and -2 and the p-value threshold of 0.05. Also color up just the genes that meet both these thresholds. These are the ones we will focus on next day! 

```{r}
mycols <- rep("gray", nrow(res))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "red" 

inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "blue"

plot( res$log2FoldChange,  -log(res$padj), 
 col=mycols, ylab="-Log(P-value)", xlab="Log2(FoldChange)" )

abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```
```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), 
                     keytype="ENSEMBL",        
                     column="SYMBOL",          
                     multiVals="first")
```

```{r}
library(EnhancedVolcano)
```

```{r}
x <- as.data.frame(res)

EnhancedVolcano(x,
    lab = x$symbol,
    x = 'log2FoldChange',
    y = 'pvalue')
```

## Save our results to date

```{r}
write.csv(res,file="myresults.csv")
```

## Adding annotation data

```{r}
head(res)
```

> Q11. Run the mapIds() function two more times to add the Entrez ID and UniProt accession and GENENAME as new columns called `res$entrez`, `res$uniprot` and `res$genename`.

```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="ENTREZID",
                     keytype="ENSEMBL",
                     multiVals="first")

res$uniprot <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="UNIPROT",
                     keytype="ENSEMBL",
                     multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res),
                     column="GENENAME",
                     keytype="ENSEMBL",
                     multiVals="first")

head(res)
```

```{r}
ord <- order( res$padj )
#View(res[ord,])
head(res[ord,])
```

```{r}
write.csv(res[ord,], "deseq_results.csv")
```


