---
title: "Class 7: Machine Learning 1"
author: "Mason Dinh (PID: A19140455)"
format: pdf
toc: true
---

## Background

Today we will explore some core machine learning methods that are very popular in bioinformatics. These include **clustering** and **dimensionallity reduction**.


## K-means clustering

The main function in "base" R for K-means clustering is called `kmeans()`

Before we go too deep let's make up some "simple" data that we can cluster and know if we are getting a good answer or not. To do this we can use `rnorm()` function:

```{r}
rnorm(100)
```

```{r}
hist(rnorm(10000,mean = 3, sd=1))
```

```{r}
x<- c(rnorm(30,-3), rnorm(30,+3))
x
```

```{r}
rev(x)
```

```{r}
z<-cbind(x = x, y = rev(x))
z
```

```{r}
p <- 1:5
cbind(p, rev(p))
```

Now we can run `kmeans()` on this input `z` and see what the results look like. 

```{r}
km<-kmeans(z, centers = 2)
km
```
```{r}
attributes(km)
```

> Q. how many points are in each cluster?

```{r}
km$size
```

> Q. What "components of your result object details 
cluster assignment/membership? 

```{r}
km$cluster
```


> Q. What "components of your result object details cluster center?

```{r}
km$centers
```

> Q. plot `z` colored by the kmeans cluster assignment and add cluster centers as blue points.

```{r}
plot(z,col=km$cluster)
points(km$centers,col="blue",pch=15)
```

> Q. Run a K-means clustering and plot the results asking for 4 clusters (K=4)? 

```{r}
km4 <- kmeans(z, centers =4 )
plot(z, col = km4$cluster)
points(km$centers,col="black",pch=15)
```

> **N.B.** you need to tell K-means the number of clusters (i.e. set `centers=2`)!!

```{r}
km$totss
```

One approach is to try different values for `centers` and then pick the best...

```{r}
ans<- NULL
for(i in 1:10) {
km <- kmeans(z,centers=i)
ans<-c(ans,km$tot.withinss)
}
ans
plot(ans, typ="o", xlab= "Number of Clusters", ylab="Total Sum of Squars Distance")
```

## Hierarchial Clustering

The main function in "base" R for hierarchical clustering is called `hclust()`

This function does not take your "raw" data for clustering. You must first build a "distance matrix" from your data and pass this as input to `hclust()`

```{r}
d <- dist(z)
hc <- hclust(d)
hc
```

There is a bespoke `plot()` method for `hclust()` result objects. 

```{r}
plot(hc)
abline(h=8,col="red")
```

Once we have our `hclust()` object (our "tree" of "cluster dendrogram") we can *"cut"* the tree to reval the clustering pattern.

```{r}
cutree(hc,h=8)
```

```{r}
cutree(hc, k=4)
```

> Q. Make a plot of `z` with your hclust results (i.e. colored by cluster membership)

```{r}
grps<-cutree(hc,k=2)
plot(z, col=grps)
```

## Principal Component Analysis (PCA)

PCA is a dimensionality reduction method that is popular for revealing patterns in complex datasets 

## Analysis of UK food data

Let's look at some data on the eating habits of folks from the UK to see if there are patterns and trends that have some regions being distinct from others. 

## Data Import

The data is made available in CSV format so we can use the `read.csv()` function

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```
> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

```{r}
head(x)
```

```{r}
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

```{r}
dim(x)
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?

I prefer the second method to solving the "row-names problem" because it is much more convenient and can be done much faster than the first method. However, I think the first method is much more robust because if gives the user a lot more control over the output of the code. 

```{r}
library(tidyr)
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")
dim(x_long)
```

```{r}
library(ggplot2)
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "dodge") +
  theme_bw()
```

> Q4: Changing what optional argument in the above ggplot() code results in a stacked barplot figure?

Changing the position from "dodge" to "stack" will result in a stacked barplot.

```{r}
ggplot(x_long) +
  aes(x = Country, y = Consumption, fill = Food) +
  geom_col(position = "stack") +
  theme_bw()
```

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

The pairs() function creates scatterplots that show the relationship between every pair of variables in the dataset. If a point lies on a diagonal line in the figure, it means the two variables that are being compared have the same value. 

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

```{r}
library(pheatmap)

pheatmap( as.matrix(x) )
```
> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?

The countries that cluster together are England, Wales, and Scotland because from the plots we can see that their points follow similar trends. This suggest that these three countries share similar food consumption patterns. You can tell that there is a difference between Northern Ireland and the other countries but its hard to see anything more specific.

## Tidy data

```{r}
library(tidyr)
x_long <- x |> 
          tibble::rownames_to_column("Food") |> 
          pivot_longer(cols = -Food, 
                       names_to = "Country", 
                       values_to = "Consumption")
dim(x_long)
```

```{r}
head(x_long)
```

Fix anything that went wrong with data import.

## Exporatory analysis
Make some plots to help make sense of obvious trends...

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

> Q3: Changing what optional argument in the above barplot() function results in the following plot?

Changing the `beside()` function from TRUE to FALSE will stack the bars.

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))

```

> **Key-point**: Even relatively small datasets can prove challenging to interpret

## PCA to the rescue

The main function in "base" R for PCA is called `prcomp()`. This function wants the "observations" to be rows and the "variables to be columns.

So here we need to take the transpose of our `x` input object

```{r}
pca <- prcomp(t(x))
summary(pca)
```

The returned `pca` object has components that we can use to make our main result figures:

```{r}
attributes(pca)
```

The main result figure from this analysis is called a "PC score plot" or "ordination plot" or "PC plot" or PC1 vs PC2 plot. 

> Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.

```{r}
library(ggplot2)
ggplot(pca$x) + aes(PC1,PC2) + geom_point()
```

```{r}
mycols<-c("orange","red","blue","darkgreen")

library(ggplot2)
ggplot(pca$x) + aes(PC1,PC2) + geom_point(col=mycols)
```

> Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document.

```{r}
mycols<-c("orange","red","blue","darkgreen")

library(ggplot2)
ggplot(pca$x) + aes(PC1,PC2,label=row.names(pca$x)) + geom_point(col=mycols) +  geom_text(size = 3, vjust = 2, col=mycols)
```

> Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?

The two food groups are soft drinks and fresh potatoes. PC2 mainly tells us about the variables that distinguish certain regions. PC2 tells us about the contrast between the food type and the country.

```{r}
ggplot(pca$rotation)+ aes(PC1, row.names(pca$rotation)) + geom_col()
  
```





